Hydrogen Epoch of Reionization Array (HERA)

### 2016-04-18 manuscript.tex
editing and restructure of sec 3

\subsection{The ``Wedge''}
\label{sec:wedge}
Perhaps the most important advance informing HERA's design is a
refined understanding of how smooth-spectrum foregrounds interact
with instrument chromaticity to produce a characteristic ``wedge" of
foreground leakage in Fourier space (see Fig.~\ref{fig:wedge}), 
outside of which the 21\,cm signal dominates (the ``EOR window'').  The coordinates of the wedge are discussed below.
Through theoretical and observational work
\citep{datta_etal2010,morales_et_al2012,parsons_et_al2012b,vedantham_2012,thyagarajan_et_al2013,hazelton_et_al2013,pober_etal2013b,liu_et_al2014a,liu_et_al2014b},
we have learned how the boundary between the wedge and  the EoR window is determined by the separation between antennas,
signal reflections within antennas, and the angular response of the antenna beam.  Deep integrations also show us
that, to the limits of current sensitivity, foreground emission is absent outside of the wedge; it can only 
appear there through instrumental leakage \citep{parsons_etal2014,ali_et_al2015,moore_et_al2016,kohn_et_al2016}.

Though the power spectrum measurement is over a volume, it is instructive to split the spatial wavenumbers into two components,  $\kvec =  \kvpr + \kpar{\hat {\bf z}}$ where $|\kvpr|\equiv \kpr = 2\pi b/\lambda X$ is determined by the antenna baseline ($b$) and $\kpar=2\pi/Y\Delta\nu$ by the bandwidth ($\Delta\nu$).  Here $X$ and $Y$ are cosmological parameters relating angular size and spectral frequency to cosmic volumes respectively (so, relating wavenumber to physical volume at a given redshift).  This is useful since it allows us to split out the chromatic response of the interferometer visibility measurement and isolate a phase space where foreground sources ({\em i.e.} everything {\em not} the EOR) contaminate the signal of interest from where they don't.

In these variables, the ``contaminated'' phase space is a wedge-shaped region in $\kpr$:$\kpar$ space such that the ratio of the time-delay across a given baseline ($b/c$,  which impacts the chromatic visibility) and the delay associated with a given bandwidth ($1/\Delta\nu$) is less than some parameter determined by the details of the system, which we denote as $1/\beta$.  Substituting $\kpr$ and $\kpar$ in for $b$ and $\Delta\nu$ in this ratio, this wedge is bounded by
\begin{equation}
\kpar \le \beta\frac{X\lambda}{Yc}\kpr + \frac{S}{Y}
\end{equation}
where an offset parameterized by $S$ accounting for affects related to the combined spectral smoothness of the foregrounds and the antenna response has been included(see Fig. \ref{fig:wedge}).

@@

The techniques to measure the power spectrum may be broken down into two principle techniques -- delay-space (\ref{sec:delayapproach}) and map-making (\ref{sec:mapapproach}) -- plus additional hydrid methods.  These are briefly summarized below, with an emphasis on the delay-spectrum technique as the initial approach taken by HERA.  Finally, we provide a short discussion on calibration in the concept of this design.

@@

Fig. \ref{fig:boost} (left) plots this ``redundancy boost'' factor for various configuration types relative to an array in a 19-element hexagonal configuration.  They are labelled for the type of configuration (hex, grid, imaging) and the number of antennas (from 37 to 350).  Note that (a) {\em hex} is based on the configuration used for HERA (14.6m center-to-center spacing); (b) {\em grid} for that of PAPER (4m N-S, 16m E-W spacing); (c) {\em imaging-128} for MWA \citep{tingay_et_al2013}; and (d) {\em imaging-48} for LOFAR (\citealt{2013A&A...556A...2V}; note that this assumes the ability to fully correlate the HBA-pairs).  The plot shows that for a fixed number of elements using the delay-spectrum approach, redundant arrays (hex/grid) provide about an order of magnitude improvement over imaging arrays.  Since the baselines go as order $N^2$, this corresponds to using about 1/3 of the number of elements to yield the same performance (as verified by the Hex-37/Imaging-128 lines).  This advantage does not extend to the other approaches discussed below and imaging configurations would be preferred.  Note that a filled hexagonal-packed array provides excellent imaging as well, but with potentially limited resolution unless outriggers are included.

Another approach to improving sensitivity is to make the collecting area per element larger.  In the limit, this yields a fully filled aperture as mentioned.  A larger element limits the accessible field-of-view, which however is not a huge liability for a focussed experiment, as long as an appropriate patch of ``cold'' sky passes within its field-of-view.  One method combining both advantages is to use small elements to build up phased array ``tiles'' and then electronically produce many beams, which may get correlated.  This adds additional complexity and potential systematics, however it is the method employed by both MWA and LOFAR.  Fig. \ref{fig:boost} (right) shows the relative sensitivity of the various array relative to HERA-19, calculated for a redshift corresponding to the peak signal at a hydrogen reionization fraction of 0.5.  The circles near the array name indicate the relative sizes of the element.  Note that MWA and LOFAR are cross-hatched to indicated phased array tiles.

@@

The power spectrum statistic uses the full magnitude of the {\bf k}-vector, which is set by both the antenna baseline and bandwidth, as discussed above.  The $\kvpr$ components are proportional to the visibilities and $\kpar$ are proportional to the Fourier transform of the frequency response.   The Fourier transform of a frequency spectrum is a delay, hence the name ``delay-spectrum'' for this technique.  Note that cosmic evolution limits the largest bandwidth (which determines the smallest $\kpar$) to about 10 MHz -- for larger bandwidths the evolution of the Universe begins to impact the result.  For HERA, wavenumbers are dominated by the bandwidth, not the baseline.

@@

Figure \ref{fig:kperf} shows many of the dependencies on wavenumber of bandwidth, configuration and the cosmos.  It shows the perpendicular, parallel and total resolution wavenumber as a function of redshift for various bandwidths and baselines.  The term {\em resolution} wavenumber is used, since it is a lower limit for that parameter since larger wavenumbers may be obtained by channelizing.  The redshift range is appropriate for the extended frequency range that is the goal for HERA, however $z$=6-13 is the primary region for EOR studies.

In Fig. \ref{fig:kperf}, blue lines denote $\kpar$ values, black lines denote $\kpr$, and green lines denote total magnitude ($|{\kvec}|$).  The lines are labelled for the assumed bandwidth and/or baseline, depending on the type of wavenumber shown.  Obviously many values may be used, but here a small sampling of values appropriate for HERA are shown.    Also shown just below the blue $\kpar$ curves are redshift ranges for that bandwidth centered at $z$=6,12, and 18 (they appear as horizontal ``error'' bars).  Note that for 100 MHz, the concept of a power spectrum at a given redshift is no longer appropriate, so some evolutionary model would have to be assumed (this is why the line is dashed and only the $z$=12 range is shown).
Also in Fig. \ref{fig:kperf}, the lime-green shaded region indicates a plausible region in which HERA could operate and the lighter square within that region is the area that is being targeted in the first phase.

\begin{figure}[h!]
\centerline{
\includegraphics[width=.4\textwidth]{plots/kperf.pdf} 
}
\caption{\small $|\kvec |, \kpar, \kperp$ wavenumber plots at various bandwidths ($k_{||}$; blue), baselines ($\kperp$; black), and total ($|\kvec |$, green) as a function of redshift.  The stepped profiles for $k_{||}$ and $|\kvec |$ are the line-of-site redshift ranges for that bandwidth (too small to see at 1 MHz).  See text for discussion.  The full redshift range is the HERA extended goal.  The initial HERA delay-spectrum power spectrum goal is the light square region in the upper left.}
\label{fig:kperf}
\end{figure}

\subsection{Mapmaking Approach}
\label{sec:mapapproach}
In contrast to the delay-spectrum approach, mapmaking approaches combine baselines to build up information before squaring and calculating the power spectrum.  The conceptually simpler approach is to first make an image cube of the sky as a function of angular position and frequency, and then to Fourier transform, square, and bin the cube to determine the power spectrum \cite{liu_tegmark2011,dillon_et_al2013a}. The image domain is a natural place to combine information from partially-coherent pairs of visibility measurements (either from different baselines or from the same baseline at different times). Mapmaking can be lossless data compression step \citep{tegmark1997b}, though keeping track of statistical properties of the maps, which have complex frequency and position dependent point spread functions and noise covariance matrices, can be computationally challenging \citep{dillon_et_al2015a}. Existing approaches have had to make a number of approximations, including that point spread functions do not vary over the field of view and that noise is not correlated between $uv$-cells. Additionally, mapmaking is in principle not as vulnerable to polarization leakage as the delay-spectrum approach, since only polarization mismodeling (rather than any kind of polarization asymmetry) can cause leakage from Stokes $Q$ or $U$ to $I$.

@@

While they were presented separately above, the delay-spectrum and map-making approaches need not be viewed as mutually exclusive. For example, mapmaking need not be limited to the real space image basis; pipelines that combine multiple baselines into estimates of various Fourier amplitudes of the sky are formally also mapmaking pipelines. Like the delay-spectrum approach, such pipelines (see, e.g., \citealt{trott_et_al2016}) avoid the image domain entirely. This helps to prevent artifacts that may introduced by imaging algorithms.

@@

Fundamentally, the hybrid approach and the optimal mapmaking approach \citep{dillon_et_al2015a} are very similar. Whether it makes computational sense to examine all pairs of partially-coherent visibilities or to perform mapmaking as an intermediate data-compression step before power spectrum estimation (despite the complexity of the resultant map statistics) is an open question that depends on a number of factors, including the number of elements in the array, its layout, and its integration time.
These approaches need further development, and can be incorporated when sufficiently implemented.

@@

However, at each frequency, degeneracies in the redundant calibration procedure prevent it from solving four numbers per frequency for the whole array (down from 700 in HERA's case). The first is an overall gain---one can multiply each $g_i(\nu)$ by 2 and divide each $V_{ij}^\text{true}$ by 4, for example, and get exactly the same measured visibilities. The others are an overall phase and two other phase terms that describe the tip and tilt of the array. These degeneracies mean that the output of any redundant baseline calibration also need to undergo a final, ``absolute'' calibration. Absolute bandpass calibration has thus far been performed for PAPER and MITEoR using a sky model, though it remains an open question whether this reduced set of calibration parameters can be further constrained without an explicit sky model.

>>>>

\subsection{The ``Wedge''}
\label{sec:wedge}
Perhaps the most important advance informing HERA's design is a
refined understanding of how smooth-spectrum foregrounds interact
with instrument chromaticity to produce a characteristic ``wedge" of
foreground leakage in Fourier space (see Fig.~\ref{fig:wedge}), 
outside of which the 21\,cm signal dominates (the ``EOR window'').  
Through theoretical and observational work
\citep{datta_etal2010,morales_et_al2012,parsons_et_al2012b,vedantham_2012,thyagarajan_et_al2013,hazelton_et_al2013,pober_etal2013b,liu_et_al2014a,liu_et_al2014b},
we have learned how the boundary between the wedge and  the EoR window is determined by the separation between antennas,
signal reflections within antennas, and the angular response of the antenna beam.  Deep integrations also show us
that, to the limits of current sensitivity, foreground emission is absent outside of the wedge; it can only 
appear there through instrumental leakage \citep{parsons_etal2014,ali_et_al2015,moore_et_al2016,kohn_et_al2016}.

The power spectrum measurement provides the spatial correlations across the sky, characterized by the magnitude of the wavenumber, ${\bf k}$.
Though the full magnitude of the {\bf k}-vector is used, it is instructive to split it into two components,  $\kvec =  \kvpr + \kpar{\hat {\bf z}}$ where $|\kvpr|\equiv \kpr = 2\pi b/\lambda X$ is determined by the antenna baseline ($b$) and $\kpar=2\pi/(YB)$ by the bandwidth ($B$).  Here $X$ and $Y$ are cosmological parameters relating angular size and spectral frequency to cosmic volumes respectively (so, relating wavenumber to physical volume at a given redshift).  This is useful since it allows us to split out the chromatic response of the interferometer visibility measurement and isolate a phase space where foreground sources ({\em i.e.} everything {\em not} the EOR) contaminate the signal of interest from where they don't.

The $\kvpr$ components are directly proportional to the baselines and $\kpar$ are proportional to the Fourier transform of the frequency response.   The Fourier transform of a frequency spectrum is a delay, hence the name ``delay-spectrum'' for the technique that uses this approach.  Note that cosmic evolution limits the largest bandwidth (which determines the smallest $\kpar$) to about 10 MHz -- for larger bandwidths the evolution of the Universe begins to impact the result.  For HERA, wavenumbers are dominated by the bandwidth, not the baseline.

Figure \ref{fig:kperf} shows many of the dependencies on wavenumber by the bandwidth, configuration and cosmos by plotting the perpendicular wavenumber ($\kvpr$; black lines), parallel wavenumber ($\kpar$; blue lines) and total wavenumber ($k$; green lines) as a function of redshift for various bandwidths and baselines.  The redshift range is appropriate for the extended frequency range that is the goal for HERA, however $z$=6-13 is the primary region for EOR studies.
The lines are labelled for the assumed bandwidth and/or baseline, depending on the type of wavenumber shown.  Obviously many values may be used, but here a small sampling of values appropriate for HERA are shown.    
Wavenumbers with an assumed bandwidth ($k$ and $\kpar$ are shown as stepped profiles, with the redshift steps equal to the bandwidth at that redshift.
Note that for 100 MHz, the concept of a power spectrum at a given redshift is no longer appropriate, which is why the line is dashed.
Also in Fig. \ref{fig:kperf}, the lime-green shaded area indicates the initial region in which HERA intends to detect and characterize the EOR using the delay-spectrum technique.

\begin{figure}[h!]
\centerline{
\includegraphics[width=.4\textwidth]{plots/kperf.pdf} 
}
\caption{\small $|\kvec |, \kpar, \kperp$ wavenumber plots at various bandwidths ($k_{||}$; blue), baselines ($\kperp$; black), and total ($|\kvec |$, green) as a function of redshift.  The stepped profiles for $k_{||}$ and $|\kvec |$ are the line-of-site redshift ranges for that bandwidth (too small to see at 1 MHz).  See text for discussion.  The full redshift range is the HERA extended goal.  The initial HERA delay-spectrum power spectrum goal is the light green square region in the upper left.}
\label{fig:kperf}
\end{figure}

In these variables, the ``contaminated'' phase space is a wedge-shaped region in $\kpr$:$\kpar$ space such that the ratio of the time-delay across a given baseline ($b/c$,  which impacts the chromatic visibility) and the delay associated with a given bandwidth ($1/B$) is less than some parameter determined by the details of the system, which we denote as $1/\beta$.  Substituting $\kpr$ and $\kpar$ in for $b$ and $\Delta\nu$ in this ratio, this wedge is bounded by
\begin{equation}
\kpar \le \beta\frac{X\lambda}{Yc}\kpr + \frac{S}{Y}
\end{equation}
where an offset parameterized by $S$ accounting for affects related to the combined spectral smoothness of the foregrounds and the antenna response has been included(see Fig. \ref{fig:wedge}).

@@

The techniques to measure the power spectrum may be broken down into two principle techniques -- delay-space (\S\ref{sec:delayapproach}) and map-making (\S\ref{sec:mapapproach}) -- plus additional hydrid methods.  These are briefly summarized below, with an emphasis on the delay-spectrum technique as the initial approach taken by HERA.  Finally, we provide a short discussion on calibration in the concept of this design.

@@

Fig. \ref{fig:boost} (left) plots this ``redundancy boost'' factor for various configuration types relative to an array in a 19-element hexagonal configuration.  They are labelled for the type of configuration (hex, grid, imaging) and the number of antennas (from 37 to 350).  Note that (a) {\em hex} is based on the configuration used for HERA (14.6m center-to-center spacing); (b) {\em grid} for that of PAPER (4m N-S, 16m E-W spacing); (c) {\em imaging-128} for MWA \citep{tingay_et_al2013}; and (d) {\em imaging-48} for LOFAR (\citealt{2013A&A...556A...2V}; note that this assumes the ability to fully correlate the HBA-pairs).  The plot shows that for a fixed number of elements using the delay-spectrum approach, redundant arrays (hex and grid) provide about an order of magnitude improvement over imaging arrays.  Since the baselines go as order $N^2$, this corresponds to using about 1/3 of the number of elements to yield the same performance (as verified by the Hex-37/Imaging-128 lines).  This advantage does not extend to the other approaches discussed below and imaging configurations would be preferred.  Note that a filled hexagonal-packed array provides excellent imaging as well, but with potentially limited resolution unless outriggers are included.

Another approach to improving sensitivity is to make the collecting area per element larger.  In the limit, this yields a fully filled aperture.  A larger element does limit the accessible field-of-view, but this is not a huge liability for a focussed experiment as long as an appropriate patch of ``cold'' sky passes within its field-of-view.   Fig. \ref{fig:boost} (right) shows the relative sensitivity incorporating configuration and element size of various arrays relative to HERA-19, calculated for a redshift corresponding to the peak signal at a hydrogen reionization fraction of 0.5.  The circles to the right indicate the relative sizes of the element.  Note that MWA and LOFAR are cross-hatched to indicate phased array tiles, which uses small antennas to phase up a full element response.

@@

@@

\subsection{Mapmaking Approach}
\label{sec:mapapproach}
In contrast to the delay-spectrum approach, mapmaking approaches combine baselines to build up information before squaring and calculating the power spectrum.  The conceptually simplest approach is to first make an image cube of the sky as a function of angular position and frequency, and then to Fourier transform, square, and bin the cube to determine the power spectrum \citep{liu_tegmark2011,dillon_et_al2013a}. The image domain is a natural place to combine information from partially-coherent pairs of visibility measurements (either from different baselines or from the same baseline at different times). Mapmaking can be a lossless data compression step \citep{tegmark1997b}, though keeping track of statistical properties of the maps, which have complex frequency and position dependent point spread functions and noise covariance matrices, can be computationally challenging \citep{dillon_et_al2015a}. Existing approaches have had to make a number of approximations, including that point spread functions do not vary over the field of view and that noise is not correlated between $uv$-cells. Additionally, mapmaking is in principle not as vulnerable to polarization leakage as the delay-spectrum approach, since only polarization mismodeling (rather than any kind of polarization asymmetry) can cause leakage from Stokes $Q$ or $U$ to $I$.

@@

While they were presented separately above, the delay-spectrum and map-making approaches need not be viewed as mutually exclusive. For example, mapmaking need not be limited to the real space image basis; pipelines that combine multiple baselines into estimates of various Fourier amplitudes of the sky are formally also mapmaking pipelines. Like the delay-spectrum approach, such pipelines (see, e.g., \citealt{trott_et_al2016}) avoid the image domain entirely. This helps to prevent artifacts that may be introduced by imaging algorithms.

@@

Fundamentally, the hybrid approach and the optimal mapmaking approach \citep{dillon_et_al2015a} are very similar. Whether it makes computational sense to examine all pairs of partially-coherent visibilities or to perform mapmaking as an intermediate data-compression step before power spectrum estimation (despite the complexity of the resultant map statistics) is an open question that depends on a number of factors, including the number of elements in the array, its layout, and its integration time.
However, similar to map-making techniques, these approaches need additional precision in calibration and modeling development in order to be implemented with confidence.

@@

However, at each frequency, degeneracies in the redundant calibration procedure prevent it from solving four numbers per frequency for the whole array (down from 700 in HERA's case if full calibration, as opposed to redundant calibration, were needed). The first is an overall gain---one can multiply each $g_i(\nu)$ by 2 and divide each $V_{ij}^\text{true}$ by 4, for example, and get exactly the same measured visibilities. The others are an overall phase and two other phase terms that describe the tip and tilt of the array. These degeneracies mean that the output of any redundant baseline calibration also need to undergo a final, ``absolute'' calibration. Absolute bandpass calibration has thus far been performed for PAPER and MITEoR using a sky model, though it remains an open question whether this reduced set of calibration parameters can be further constrained without an explicit sky model.
